Sulfur Dioxide on Water's Surface

### 2011-06-23 EquilibratedOrientation.tex
Draft 6

The orientation depth-profiles for \wat~are shown in figure \ref{fig:water-orientation} for both the neat-water (top) and saturated (bottom) systems during the equilibrium MD simulations. In both systems the strongest orientational preference is found at the slab surfaces (distances greater than 0\angs) where the water is furthest towards the gas phase. Previous work on orientational preference of water at air surfaces shows the same trend as our neat-water results.\cite{Walker2006b,Hore2008} The narrow "penninsula" of high intensity that extends in the $\cos(\theta)$ plots to the right of the -5\angs~location shows the overall preference of water to orient at the surface. The $\cos(\phi)$ plots are similar to each other with a narrow region of reorientation, but the effect of the surface is greater in the neat-water system as evidenced by the sharper transition in intensity from blue to red, compared to the saturated system that has a less pronounced intensity change.

@@

Within 0-5\angs~of the water surface (the approximate distance from the surface that \suldiox~begins interacting with the outer-most waters) the angular distribution of the single \suldiox~(in the neat-water system) is concentrated primarily in $\cos(\theta)>0$. The peak of the distribution occurs at $\cos(\theta)=1$. This indicates that the \suldiox~bisector points out of the water surface, with the sulfur atom pointing towards the aqueous bulk, and the two oxygens pointing into the gas phase. This same distribution occurs in the saturated system for positions below 5\angs. Beyond 5\angs~above the surface both distributions become isotropic. Promixity to the water highly orients the \suldiox~with the sulfur atom pointing in towards the water bulk. Moving further away from the water surface, and interacting less with \wat~molcules allows for greater orientational freedom as exhibited in the isotropy of the distributions above 5\angs. %Because both the neat-water and saturated systems show rather similar distributions, it is possible that the concentration of \suldiox~does not strongly affect the orientation of the \suldiox, unlike the orientations of waters.

>>>>

The orientation depth-profiles for \wat~are shown in figure \ref{fig:water-orientation} for both the neat-water (top) and saturated (bottom) systems during the equilibrium MD simulations. In both systems the strongest orientational preference is found at the slab surfaces (distances greater than 0\angs) where the water is furthest towards the gas phase. Previous work on orientational preference of water at air surfaces shows the same trend as our neat-water results.\cite{Walker2006b,Hore2008} The narrow "peninsula" of high intensity that extends in the $\cos(\theta)$ plots to the right of the -5\angs~location shows the overall preference of water to orient at the surface. The $\cos(\phi)$ plots are similar to each other with a narrow region of reorientation, but the effect of the surface is greater in the neat-water system as evidenced by the sharper transition in intensity from blue to red, compared to the saturated system that has a less pronounced intensity change.

@@

Within 0-5\angs~of the water surface (the approximate distance from the surface that \suldiox~begins interacting with the outer-most waters) the angular distribution of the single \suldiox~(in the neat-water system) is concentrated primarily in $\cos(\theta)>0$. The peak of the distribution occurs at $\cos(\theta)=1$. This indicates that the \suldiox~bisector points out of the water surface, with the sulfur atom pointing towards the aqueous bulk, and the two oxygens pointing into the gas phase. This same distribution occurs in the saturated system for positions below 5\angs. Beyond 5\angs~above the surface both distributions become isotropic. Promixity to the water highly orients the \suldiox~with the sulfur atom pointing in towards the water bulk. Moving further away from the water surface, and interacting less with \wat~molecules allows for greater orientational freedom as exhibited in the isotropy of the distributions above 5\angs. %Because both the neat-water and saturated systems show rather similar distributions, it is possible that the concentration of \suldiox~does not strongly affect the orientation of the \suldiox, unlike the orientations of waters.

### 2011-06-23 Introduction.tex
Draft 6

The doorway to the uptake of a gas by an aqueous solution is the water surface. Although we know much about the behavior of a gas on either side of that entrance, far less is known about how that surface acts to attract, facilitate, or thwart the transit of the gas between the two bulk phases. What species form during gas adsorption onto liquid surfaces, and what are the intermediary steps?  To what extent does an adsorbing gas affect liquid water surface molecules, or the structure of the interface throughout an adsorption process?  Are specific gas or liquid molecular orientations necessary for gaseous adsorption?  Experimental studies to address such questions are valuable but do not provide the resolution necessary to determine the geometries of adsorbing gases, or to determine the orientations of the molecules at the liquid surface near the adsorption site.
This type of information can be determined computationally, and when coupled to the experimental studies can provide important new insight into the gas-liquid suface adsorption process.

@@

Our recent vibational sum frequency spectroscopy (VSFS) experiments have provided insight into the adsorption and reaction of gases in atmospheric aerosols.\cite{Tarbuck2005,Tarbuck2006} These studies showed that an \suldiox~surface hydrate complex forms when an aqueous surface is exposed to \suldiox~gas. A computational study by Baer et al.\cite{Baer2010} then made a series of predictions of the specific nature of the hydrated complex through classical and ab initio simulations. That work developed a detailed picture of the nature of the \suldiox~surface complex with water, and related it to the surface water OH vibrational IR spectra.

@@

We provide such detailed informatino to complement and augment these experimental studies. Our studies involve computation of the orientational depth profiles of water and \suldiox~before and after the adsorption process. They augment the spectroscopic experiments that can not probe depth or orientation profiles of surface species in the same detailed manner as computational simulations. The molecular behaviors that occur upon introduction of \suldiox~to a \wat~surface are determined, and we analyze the orientational response of both the \suldiox~and \wat~molecules using equilibrium and steered (SMD) classical molecular dynamics simulations. The latter approach involves steering a gas molecule into the aqueous phase, and characterizing its molecular orientation as it transits through the interfacial region.  This unique approach enables new insights into the behaviors of gas molecules as they move near to liquid water to be obtained. We also simulate how \suldiox~adsorption occurs on a water surface saturated with adsorbed \suldiox, analogous to the cold \suldiox~experimental studies recently performed.\cite{Ota2011} The results of this study provide an intimate perspective on the adsorption of \suldiox~at an aqueous surface, and a more complete picture of gaseous adsorption to liquid interfaces.

>>>>

The doorway to the uptake of a gas by an aqueous solution is the water surface. Although we know much about the behavior of a gas on either side of that entrance, far less is known about how that surface acts to attract, facilitate, or thwart the transit of the gas between the two bulk phases. What species form during gas adsorption onto liquid surfaces, and what are the intermediary steps?  To what extent does an adsorbing gas affect liquid water surface molecules, or the structure of the interface throughout an adsorption process?  Are specific gas or liquid molecular orientations necessary for gaseous adsorption?  Experimental studies to address such questions are valuable but do not provide the resolution necessary to determine the geometries of adsorbing gases, or to determine the orientations of the molecules at the liquid surface near the adsorption site.
This type of information can be determined computationally, and when coupled to the experimental studies can provide important new insight into the gas-liquid surface adsorption process.

@@

Our recent vibrational sum frequency spectroscopy (VSFS) experiments have provided insight into the adsorption and reaction of gases in atmospheric aerosols.\cite{Tarbuck2005,Tarbuck2006} These studies showed that an \suldiox~surface hydrate complex forms when an aqueous surface is exposed to \suldiox~gas. A computational study by Baer et al.\cite{Baer2010} then made a series of predictions of the specific nature of the hydrated complex through classical and ab initio simulations. That work developed a detailed picture of the nature of the \suldiox~surface complex with water, and related it to the surface water OH vibrational IR spectra.

@@

We provide such detailed information to complement and augment these experimental studies. Our studies involve computation of the orientational depth profiles of water and \suldiox~before and after the adsorption process. They augment the spectroscopic experiments that can not probe depth or orientation profiles of surface species in the same detailed manner as computational simulations. The molecular behaviors that occur upon introduction of \suldiox~to a \wat~surface are determined, and we analyze the orientational response of both the \suldiox~and \wat~molecules using equilibrium and steered (SMD) classical molecular dynamics simulations. The latter approach involves steering a gas molecule into the aqueous phase, and characterizing its molecular orientation as it transits through the interfacial region.  This unique approach enables new insights into the behaviors of gas molecules as they move near to liquid water to be obtained. We also simulate how \suldiox~adsorption occurs on a water surface saturated with adsorbed \suldiox, analogous to the cold \suldiox~experimental studies recently performed.\cite{Ota2011} The results of this study provide an intimate perspective on the adsorption of \suldiox~at an aqueous surface, and a more complete picture of gaseous adsorption to liquid interfaces.

### 2011-06-23 Methods.tex
Draft 6

All simulations began with an equilibrated cube of 900 \wat~molecules, with sides of length 30\angs. The long axis of each simulation cell (the axis normal to the water surface) was then lengthened to 120\angs, and the systems were further equilibrated for 10 ns. The simulations all employed periodic boundaries to create an ``infinite-slab'' geometry. After equilibrating the neat-\wat~slabs two types of systems were created by introducing \suldiox: a single-\suldiox~system, herein referred to as the ``neat-water'' system, and a saturated \suldiox~system with multiple gaseous \suldiox~molecules.

The low and high concentration simulated systems, ``neat-water'' and saturated, respectively, were created as follows: the ``neat-water'' simulation involved the addition of a single \suldiox~molecule either within the bulk of the water slab (for equilibrium MD), or above the slab surface (in the SMD simulations). The single-\suldiox~neat-water system was then evolved for 2 ns to produce an equilibrated starting configuration. Because of the extremely low concentration of \suldiox~in the neat-water system the surface waters behave similarly to a true ``neat-water'' air-liquid interface. The neat-water system orientational results shown later in this work reproduce well the results of our previous orientational studies of surface water behavior.\cite{Walker2006b,Hore2008} The saturated system had 22 \suldiox~molecules introduced to the water slab bulk in order to saturate it to a level coinciding with the Henry's law constant for \suldiox~in water. Additionally, 50 \suldiox~molecules were introduced into the gas phase outside of the saturated water slab to simulate an \suldiox~gas pressure. The saturated system was then evolved for 2 ns to produce a starting configuration for further saturated simulations.

@@

Equilibrium simulations involved adding \suldiox~to a water slab and equilibrating as outlined above. The neat-water system had a single \suldiox~added to the center of the water box, representing a concentration of 0.06 M. The more concentrated ``saturated'' system consisted of 22 \suldiox~molecules in the bulk corresponding to a concentration 1.35 M. This saturated system was exposed to an additional 50 \suldiox~in the gas phase above the water surface. Both the low and high concentration systems were then evolved for a further 10 ns using a timestep of 0.5fs, with atomic coordinates recorded every 100 fs.

@@

% need to add in the specifics about the SMD routine, force constants, calculations, etc.
A second set of simulations began with an equilibrated water slab as in the surface equilibrated method above. However, in both the equilibrated neat-water and saturated systems, a single \suldiox~was introduced 20\angs~above the water slab surface, with the sulfur atom tethered to its initial position. The systems were evolved for 1 ns, taking coordinate snapshots every 20 ps to create 50 starting points for further simulations. Steered molecular dynamics (SMD) were then performed on the 50 system configurations (in both the neat-water and saturated configurations) to guide the \suldiox~down towards a tethered water near the water slab's center of mass by applying a small steering force to the \suldiox-sulfur atom. This steering technique has been previously developed and used to successfuly model chemical events.\cite{Isralewitz2001} The \suldiox~thus passed through the continuum of environments from gas phase to (neat- and saturated) water surface adsorption, and finally absorption into the bulk of the \wat~slab. Each of the SMD simulations were performed for a total of 200 ps, using a timestep of 1 fs, and taking snapshots of the system every 25 fs. Figure \ref{fig:starting-configurations} illustrates two sample starting configurations for the SMD simulations, showing both the neat-water slab and the saturated slab configurations before steering the \suldiox~towards the water bulk.

@@

	In this portion of the study we create orientational depth-profiles of \suldiox~and \wat~upon exposure and adsorption of the gas. The reference point used for depth calculations is the location of the water surface. A liquid surface, however, is a dynamic boundary that is neither flat nor stationary. To define a water surface location, several previous studies have used the technique of fitting a lineshape to the averaged density profile of the water, and extracting interfacial shape and location parameters.\cite{Shamay2010,Wick2006c,Chowdhary2006} Hyperbolic tangent functions have been used often, and values for the ``Gibb's dividing surface'' location, and interfacial width have thus been determined.\cite{Matsumoto1988} However, in long simulations the location and shape of the interface changes, and the motion of surface waters alters the interfacial width at any given timestep. Thus, the density profile fitting will capture averaged widths and locations, not instantaneous values. Similarly, the averaged values of location and width will obscure information about any drift or deformations the surface undergoes.

Figure \ref{fig:density-flaw} demonstrates the problem of surface location drift during a simulation. Figure \ref{fig:density-flaw}a shows the density profile of water and \suldiox~over the course of one of the 10 ns trajectories used in this work using the original uncorrected coordinates of the system. The water density profile and location (dashed line in figure \ref{fig:density-flaw}a) is produced by averaging the instantaneous density profile at each timestep in the simulation over all the timesteps. The water profile was then fit to a $\tanh$ function (black line) to extract the position and width parameters of the water surface. The fitted Gibbs dividing surface has a width of 3.77\angs, which is comparable to values reported for similar neat-water systems.\cite{Dang1997,Hore2008} A bulge in the gas-phase ($\leq 52$\angs) side of the water density profile is indicative of the drift of the water slab over the course of the trajectory. Thus the calculated location and width from the $\tanh$ line fit are not accurate over long trajectories for defining a stationary reference point.

	In this work we define the surface location by calculating a reference location at each timestep by averaging the positions of the waters contained in the topmost monolayer. This provides us with a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden.

The location of the water surface is determined by the averaged positions of a set number of the outer-most waters in the simulated aqueous slabs. The number of waters included in the averaging is determined by taking a few issues into account.  First, counting the waters found in the topmost cross-section of the water slab over several timesteps indicated between 65-75 waters that established a full monolayer. This was done by a visual inspection of the slab using the VMD MD visualization package.\cite{HUMP96} Alternatively, assuming a spherical model of water with a radius of 2.2\angs~(the typical length of a bulk-water hydrogen bond), two layers of hexagonally tight-packed spheres yielded a similar number of surface water molecules. Increasing the number of waters used in calculating the surface location diminishes the effects of the few waters that briefly rise above the surface into the gas phase, stabilizing both the surface position and thickness values. Taking the close-packed model as a maximum number of waters fit into a flat surface, the authors decided to use 70 water molecules for calculating instantaneous water surface locations.

Using this alternative method for finding the outer monolayer location was implemented, and the surface location is plotted as a function of simulation time in figure \ref{fig:density-flaw}b. It is apparent from the surface location plot that the water slab location, and thus the surface location, drifts over the 10 ns, spanning approx. 12 $\AA$. However, the maximum standard deviation of the positions of the waters comprising the surface layer at each timestep is only 1.85 $\AA$. Consequently, all depth locations in our analyses are calculated relative to the instantaneous surface location at the corresponding timestep of the trajectories.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/density/density-location.png}
		\caption{(A) Density profiles of \wat~(dashed) and \suldiox~(red) from a 10 ns simulation of the neat-water system with a single sulfur dioxide. The \wat~density profile was fit using a hyperbolic tangent function (black) to extract location and width data for the water surface. (B) The instantaneous location of the outer \wat~monolayer was calculated for each simulation timestep (blue), and the surface location extracted from the density fitting was also shown for reference (horizontal dashed line). Long simulations of liquid slabs result in drift of the slab location, and the consequent broadening of the water density profile if the surface location is not determined repeatedly throughout the simulation.}
		\label{fig:density-flaw}
	\end{center}
\end{figure}

>>>>

All simulations began with an equilibrated cube of 900 \wat~molecules, with sides of length 30\angs. The long axis of each simulation cell (the axis normal to the water surface) was then lengthened to 120\angs, and the systems were further equilibrated for 10 ns. The simulations all employed periodic boundaries to create an ``infinite-slab'' geometry. After equilibrating the neat-\wat~slabs two types of systems were created by introducing \suldiox: a single-\suldiox~system, herein referred to as the ``neat-water'' system, and a saturated \suldiox~system with many gaseous surface and bulk-water \suldiox~molecules.

The low and high concentration simulated systems, ``neat-water'' and saturated, respectively, were created as follows: the ``neat-water'' simulation involved the addition of a single \suldiox~molecule either within the bulk of the water slab (for equilibrium MD), or above the slab surface (in the SMD simulations). The single-\suldiox~neat-water system was then evolved for 2 ns to produce an equilibrated starting configuration. Because of the extremely low concentration of \suldiox~in the neat-water system the surface waters behave similarly to a true ``neat-water'' air-liquid interface. The neat-water system orientational results shown later in this work reproduce well the results of our previous orientational studies of surface water behavior.\cite{Walker2006b,Hore2008} The saturated system had 22 \suldiox~molecules introduced to the water slab bulk in order to saturate it to a level coinciding with the Henry's law constant for \suldiox~in water. Additionally, 50 \suldiox~molecules were introduced into the gas phase outside of the saturated water slab to simulate an added \suldiox~gas pressure. The saturated system with both bulk and gaseous \suldiox~was then evolved for 2 ns to produce a starting configuration for further saturated simulations.

@@

Equilibrium simulations involved adding \suldiox~to a water slab and equilibrating as outlined above. The neat-water system had a single \suldiox~added to the center of the water box, representing a concentration of 0.06 M. The more concentrated ``saturated'' system consisted of 22 \suldiox~molecules in the bulk corresponding to a concentration 1.35 M. This saturated system was exposed to an additional 50 \suldiox~in the gas phase above the water surface. After equilibration for 2 ns, both the low and high concentration systems were then evolved for a further 10 ns data collection using a time step of 0.5fs, with atomic coordinates recorded every 100 fs.

@@

% need to add in the specifics about the SMD routine, force constants, calculations, etc.
A second set of simulations began with an equilibrated water slab as in the surface equilibrated method above. However, in both the equilibrated neat-water and saturated systems, a single \suldiox~was introduced 20\angs~above the water slab surface, with the sulfur atom tethered to its initial position. The systems were evolved for 1 ns, taking coordinate snapshots every 20 ps to create 50 starting points for further simulations. Steered molecular dynamics (SMD) were then performed on the 50 system configurations (in both the neat-water and saturated configurations) to guide the \suldiox~down towards a tethered water near the water slab's center of mass by applying a small steering force to the \suldiox-sulfur atom. This steering technique has been previously developed and used to successfully model chemical events.\cite{Isralewitz2001} The \suldiox~thus passed through the continuum of environments from gas phase to (neat- and saturated) water surface adsorption, and finally absorption into the bulk of the \wat~slab. Each of the SMD simulations were performed for a total of 200 ps, using a time step of 1 fs, and taking snapshots of the system every 25 fs. Figure \ref{fig:starting-configurations} illustrates two sample starting configurations for the SMD simulations, showing both the neat-water slab and the saturated slab configurations before steering the \suldiox~towards the water bulk.

@@

	In this portion of the study we create orientational depth-profiles of \suldiox~and \wat~upon exposure and adsorption of the gas. The reference point used for depth calculations is the location of the water surface. A liquid surface, however, is a dynamic boundary that is neither flat nor stationary. To define a water surface location, several previous studies have used the technique of fitting a line shape to the averaged density profile of the water, and extracting interfacial shape and location parameters.\cite{Shamay2010,Wick2006c,Chowdhary2006} Hyperbolic tangent functions have been used often, and values for the ``Gibb's dividing surface'' location, and interfacial width have thus been determined.\cite{Matsumoto1988} However, in long simulations the location and shape of the interface changes, and the motion of surface waters alters the interfacial width at any given time step. Thus, the density profile fitting will capture averaged widths and locations, not instantaneous values. Similarly, the averaged values of location and width will obscure information about any drift or deformations the surface undergoes.

Figure \ref{fig:density-flaw} demonstrates the problem of surface location drift during a simulation. Figure \ref{fig:density-flaw}a shows the density profile of water and \suldiox~over the course of one of the 10 ns trajectories used in this work using the original uncorrected coordinates of the system. The water density profile and location (dashed line in figure \ref{fig:density-flaw}a) is produced by averaging the instantaneous density profile at each time step in the simulation over all the time steps. The water profile was then fit to a $\tanh$ function (black line) to extract the position and width parameters of the water surface. The fitted Gibbs dividing surface has a width of 3.77\angs, which is comparable to values reported for similar neat-water systems.\cite{Dang1997,Hore2008} A bulge in the gas-phase ($\leq 52$\angs) side of the water density profile is indicative of the drift of the water slab over the course of the trajectory. Thus the calculated location and width from the $\tanh$ line fit are not accurate over long trajectories for defining a stationary reference point.

	In this work we define the surface location by calculating a reference location at each time step by averaging the positions of the waters contained in the topmost monolayer. This provides us with a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden.

The location of the water surface is determined by the averaged positions of a set number of the outer-most waters in the simulated aqueous slabs. The number of waters included in the averaging is determined by taking a few issues into account.  First, counting the waters found in the topmost cross-section of the water slab over several time steps indicated between 65-75 waters that established a full monolayer. This was done by a visual inspection of the slab using the VMD MD visualization package.\cite{HUMP96} Alternatively, assuming a spherical model of water with a radius of 2.2\angs~(the typical length of a bulk-water hydrogen bond), two layers of hexagonally tight-packed spheres yielded a similar number of surface water molecules. Increasing the number of waters used in calculating the surface location diminishes the effects of the few waters that briefly rise above the surface into the gas phase, stabilizing both the surface position and thickness values. Taking the close-packed model as a maximum number of waters fit into a flat surface, the authors decided to use 70 water molecules for calculating instantaneous water surface locations.

Using this alternative method for finding the outer monolayer location was implemented, and the surface location is plotted as a function of simulation time in figure \ref{fig:density-flaw}b. It is apparent from the surface location plot that the water slab location, and thus the surface location, drifts over the 10 ns, spanning approx. 12 $\AA$. However, the maximum standard deviation of the positions of the waters comprising the surface layer at each time step is only 1.85 $\AA$. Consequently, all depth locations in our analyses are calculated relative to the instantaneous surface location at the corresponding time step of the trajectories.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{images/density/density-location.png}
		\caption{(A) Density profiles of \wat~(dashed) and \suldiox~(red) from a 10 ns simulation of the neat-water system with a single sulfur dioxide. The \wat~density profile was fit using a hyperbolic tangent function (black) to extract location and width data for the water surface. (B) The instantaneous location of the outer \wat~monolayer was calculated for each simulation time step (blue), and the surface location extracted from the density fitting was also shown for reference (horizontal dashed line). Long simulations of liquid slabs result in drift of the slab location, and the consequent broadening of the water density profile if the surface location is not determined repeatedly throughout the simulation.}
		\label{fig:density-flaw}
	\end{center}
\end{figure}
